---
title: "Problem Set 5"
author: "Kerri Rose Riley"
format:
  html: default
  pdf: default
---

```{r}
# Load packages
library(tidyverse)

# Read CES cumulative data
ces <- read_dta("~/Documents/PAI741/Problem Set 5/CES Data/cumulative.dta")
```

## Question 1

Use the CCES/CES Cumulative File.

The CCES/CES cumulative dataset is used for this analysis.

## Question 2

Use only the 2024 data. Recode the five-category ideology variable into three categories: liberal, moderate, and conservative.

```{r}
# Keep only observations from 2024
ces2024 <- ces %>% filter(ces$year==2024)

# Recode ideology
ces2024$ideo5

ces2024 <- ces2024|>
  mutate(ideo3 = case_when(
    ideo5 %in% 1:2 ~ 1,
    ideo5 %in% 3 ~ 2,
    ideo5 %in% 4:5 ~ 3,
    ideo5 == 6 ~ NA
  ))

# Check recode
ces2024$ideo3

```

The new variable is coded as:

1 = Liberal  
2 = Moderate  
3 = Conservative  

## Question 3

Recode the presidential vote variable to include only the two-party vote. Respondents who voted for candidates other than Harris or Trump are coded as missing.

```{r}

colnames(ces2024)
ces2024$voted_pres_24

ces2024 <- ces2024 |>
  mutate(voted_pres_2 = case_when(
    voted_pres_24 %in% 1 ~ 1,
    voted_pres_24 %in% 2 ~ 2,
    voted_pres_24 %in% 3:8 ~ NA
  ))

ces2024$voted_pres_2

```

## Question 4

Create a table comparing ideology and presidential vote.

```{r}
# Frequency table
table(ces2024$ideo3, ces2024$voted_pres_2)

# Row proportions
prop.table(table(ces2024$ideo3, ces2024$voted_pres_2), margin = 1)

```

### What proportion of respondents who are liberal voted for Trump?

```{r}
379/(13541+379)
```

Approximately **2.7% of liberal respondents voted for Trump**.

### What proportion of respondents who are conservative voted for Harris?

```{r}
999/ (999+12907)
```

Approximately **7.2% of conservative respondents voted for Harris**.

## Question 5

Find the variable that refers to whether respondents own or rent their home. Recode "Other" as missing.

```{r}
ces2024$ownhome

ces2024 <- ces2024 |>
  mutate(clean_ownhome = case_when(
    ownhome %in% 1 ~ 1,
    ownhome %in% 2 ~ 2,
    ownhome == 3 ~ NA
  ))

ces2024$clean_ownhome
```

The new variable is coded as:

1 = Own  
2 = Rent  

## Question 6

Create a table comparing homeownership and presidential vote.

```{r}
table(ces2024$voted_pres_2, ces2024$clean_ownhome)

```

### What proportion of people who owned their homes voted for Trump? Harris?

```{r}
# Calculations for Proportion Questions
# HO Trump
14619 / (14619 + 13700)

# HO Kamala 
13700 / (14619 + 13700)
```

Approximately **51.6% of homeowners voted for Harris and 48.4% voted for Trump**.

### What proportion of people who rented voted for Trump? Harris?

```{r}
# Renters Trump
4117 / (7593 + 4117)

# Renters Harris
7593 / (7593 + 4117)
```

Approximately **35.2% of renters voted for Trump and 64.8% voted for Harris**.

## Question 7

Recode education into four categories:

1. High school degree or less
2. Some college or a two-year degree
3. Four-year college degree
4. Postgraduate education

```{r}

ces2024$educ

ces2024 <- ces2024 |>
  mutate(educ_4 = case_when(
    educ %in% 1:2 ~ 1,
    educ %in% 3:4 ~ 2,
    educ %in% 5 ~ 3,
    educ %in% 6 ~ 4
  ))

ces2024$educ_4
```

## Question 8

Create a table comparing education and presidential vote.

```{r}
table(ces2024$voted_pres_2, ces2024$educ_4)

# Column proportions
prop.table(
  table(ces2024$voted_pres_2, ces2024$educ_4),
  margin = 2
)
```

### What does education tell us about vote in the 2024 election?

In this sample, presidential vote choice varies across education categories. The proportion voting for Harris increases across the four education categories, while the proportion voting for Trump decreases. Trump received a larger share among respondents with a high school education or less, while Harris received a larger share among respondents with some college, a four-year college degree, or postgraduate education. The difference between the two candidates was largest among respondents with postgraduate education.
